Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-29 paper/gd1.tex
text thoughts

Long, thin stellar streams are also excellent laboratories for studying
small-scale structures in the mass distribution:
thin streams can remain coherent for tens of orbital periods, depending in
detail on the symmetries of the underlying galactic mass distribution (e.g.,
\citealt{Erkal:2016a}).
Encounters between massive perturbers and stream stars imprint variations in the
phase-space density that can persist and remain visible for close to their total
survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
studying or ruling out the existence of small-scale dark matter sub-halos, as
would be predicted by standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

@@

The most prominent and longest (in angular extent) thin stream is the GD-1
stream (named after its discoverers; \citealt{Grillmair:2006}).
GD-1 was found using a matched filter on photometric data from the SDSS and
was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
because of its relative proximity to the sun ($\approx 7$--$10~\textrm{kpc}$).
Despite being close and relatively high surface-density, no remnant progenitor
cluster or stellar system has been found associated with the stream, but its
width,  metallicity ($\feh \approx -1.4$; \citealt{Koposov:2010}) and estimated
stellar mass ($M \approx 2 \times 10^4~\msun$; \citealt{Koposov:2010}) are
consistent with being a disrupted globular cluster.

The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
ideal object for constraining the gravitational potential of dark matter in the
inner Milky Way.
Indeed, both from fitting orbits to binned phase-space measurements along the
GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates (\citealt{Bovy:2016}), it has been used to show that the
dark matter distribution within its orbit is consistent with being close to
spherical.

Its long length --- and therefore its implied old dynamical age --- combined
with its particular orbit also make it a prime stream to search for interactions
with dark matter subhalos:
because of its large pericentric distance and retrograde orbit with respect to
the Galactic bar, density variations in the stream are not expected from either
interactions with giant molecular clouds (\citealt{Amorisco:2016}) or resonant
encounters with the bar (\citealt{Pearson:2017}).
By estimating a stream age of $\approx 4~\textrm{Gyr}$ assuming a progenitor
mass of $10^5~\msun$, and by taking a subhalo mass function and density from
XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.

\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
(\citealt{Carlberg:2013}).
More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2017}):
[summarize their conclusions].

@@

All of the clearly identifiable portions of the GD-1 stream are located at high
Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
dust extinction or variations in extinction along the stream.
\figurename~\ref{fig:track-and-model} (top panel) shows the $V$-band extinction in the
region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
extinction map (\cite{Schlegel:1998}; hereafter SFD).
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
0.03~\textrm{mag}$.
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable.

Another \todo{segue to this paragraph...}.
However, we find that \todo{...}.
We discuss the scanning pattern in more detail in
Appendix~\ref{sec:completeness}.

@@

\todo{segue sentences...}

@@

@@

\subsection{Gaps and under-densities}
\label{sec:res_gap}

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{track-and-model.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:track-and-model}
}
\end{figure}

\subsection{Spur}
\label{sec:res_spur}

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{spur.pdf}
\end{center}
\caption{%
\todo{AB}
\label{fig:spur}
}
\end{figure}

@@

\section{Conclusions}
\label{sec:conclusions}

>>>>

Long, thin stellar streams are also excellent laboratories for studying
small-scale structures in the mass distribution:
thin streams can remain coherent for tens of orbital periods, depending in
detail on the symmetries of the underlying galactic mass distribution (e.g.,
\citealt{Erkal:2016a}).
Encounters between massive perturbers and stream stars imprint variations in the
phase-space density that can persist and remain visible for close to their total
survival time (e.g., \citealt{Yoon:2011}).
In this sense, streams represent one of the most promising directions for
testing the existence of small-scale dark matter sub-halos, predicted by 
standard $\Lambda$ Cold Dark Matter ($\Lambda$CDM) theory
(\citealt{Erkal:2015, Sanders:2016, Bovy:2017}).

@@

The most prominent and longest (in angular extent) thin stream is the GD-1
stream (named after its discoverers; \citealt{Grillmair:2006}).
GD-1 was found using a matched filter on photometric data from the SDSS and
was found to span $\approx 60^\circ$ in Heliocentric sky coordinates, partially
because of its relative proximity to the Sun ($\approx 7$--$10~\textrm{kpc}$).
Despite being close and relatively high surface-density, no remnant progenitor
cluster or stellar system has been found associated with the stream, but its
width,  metallicity ($\feh \approx -1.4$; \citealt{Koposov:2010}) and estimated
stellar mass ($M \approx 2 \times 10^4~\msun$; \citealt{Koposov:2010}) are
consistent with being a disrupted globular cluster.

The long physical length of the GD-1 stream ($\sim 15~\kpc$) and location in the
Galactic halo (pericentric distance $r_\textrm{peri} \sim 13~\kpc$) make it an
ideal object for constraining the gravitational potential of dark matter in the
inner Milky Way.
Indeed, both from fitting orbits to binned phase-space measurements along the
GD-1 stream (\citealt{Koposov:2010}) and from modeling the stream track in
phase-space coordinates \citep{Bowden:2015, Bovy:2016}, it has been used to show that the
dark matter distribution within its orbit is consistent with being close to
spherical.

Its long length --- and therefore its implied old dynamical age --- combined
with its particular orbit also make it a prime stream to search for interactions
with dark matter subhalos:
because of its large pericentric distance and retrograde orbit with respect to
the Galactic bar, density variations in the stream are not expected from either
interactions with giant molecular clouds (\citealt{Amorisco:2016}) or resonant
encounters with the bar (\citealt{Pearson:2017}).
By estimating a stream age of $\approx 4~\textrm{Gyr}$ assuming a progenitor
mass of $10^5~\msun$, and by taking a subhalo mass function and density from
XXX, \citet{Erkal:2016} predict that GD-1 could have up to one significant, wide
($\approx 5$--$7^\circ$) gap caused by an interaction with a
$10^6$--$10^7~\msun$ subhalo.
\todo{implications of subhalo detection (cite bullock review along the lines: on small scales, lcdm has some issues, so alternative classes of models proposed, some that lack structure on small scales; so ruling on the existence of subhalos expected to be detectable with GD-1 directly informs the nature of dark matter.)}

\todo{hints of density variations} from initial work Koposov:2010, later confirmed by density modeling, but all low significance
(\citealt{Carlberg:2013}).
More recently, deep photometry of GD-1 from CFHT/Megacam has revealed
interesting small-scale ``wiggles'' and significant density variations along a $\approx 45^\circ$ span of the stream (\citealt{DeBoer:2018}):
[summarize their conclusions].
\todo{these analyses relied on binned data, still contaminated by MW stars}

@@

@@

- properties of GD-1 along the track, and beyond

@@

- for study of global properties, we first find stream ridge points as a median $\phi_2$ every $x^\circ$ along $\phi_1$
- then we fit a second order polynomial to these ridge points, and define a stream track as the region $0.6^\circ$ above and beyond the best-fit polynomial (shown with two blue lines in Figure~\ref{fig:track-and-model})

- first measure density variations in background subtracted profile
- find maximum -- morphology as expected from a progenitor in the final stages of dissolution (cit)
- several prominent gaps: density contrast, width
- some (all?) seen in earlier data
- but not related to dust
All of the clearly identifiable portions of the GD-1 stream are located at high
Galactic latitudes ($b > 20^\circ$), and we therefore do not expect significant
dust extinction or variations in extinction along the stream.
\figurename~\ref{fig:track-and-model} (top panel) shows the $V$-band extinction in the
region around the GD-1 stream, computed from the Schlegel-Finkbeiner-Davis
extinction map (\cite{Schlegel:1998}; hereafter SFD).
For stream longitudes $-60^\circ < \phi_1 < 10^\circ$ and $-1^\circ < \phi_2 <
1^\circ$, the median $V$-band extinction is $\textrm{med}\left(A_V\right) =
0.04~\textrm{mag}$, and the dispersion in $A_V$ in this region is $\sigma_{A_V} \approx
0.03~\textrm{mag}$.
For longitudes $\phi_1 < -60^\circ$, as the stream approaches the Galactic disk,
dust extinction becomes more appreciable.

\begin{figure}[h]
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:track-and-model}
}
\end{figure}

- also show median properties along the stream 5deg window, 1deg spacings
- coordinates and proper motion vary
- these observations are reproduced with a simple model

% \todo{segue sentences...}

% Another \todo{segue to this paragraph...}.
% However, we find that \todo{...}.
% We discuss the scanning pattern in more detail in
% Appendix~\ref{sec:completeness}.

\subsection{Off-track features}
\label{sec:res_gap}

@@

@@

\begin{figure}[h]
\begin{center}
% \includegraphics[width=\textwidth]{spur.pdf}
\end{center}
\caption{%
\todo{APW}
\label{fig:orbit}
}
\end{figure}

- novel way to identify members -- helpful that retrograde, but hopefully also applicable to other streams as kinematic important
- gd-1 even longer -- useful for potential constraints --> future

- progenitor? if confirmed, can create more realistic models (so far only modeled the whole stream as just a leading or trailing arm)
- to confirm -- measure actions, which change direction (some bovy paper)?

- no longer simple stream, discuss possible causes
- evaporation?

- smooth component of the potential?
-- show orbit
-- chaos?
-- bar?

- encounters?
-- work out mass / time combinations in impulse approximation
-- comment on GMC vs subhalo

- future:
-- radial velocities, measure 3d velocity structure (unique predictions)
-- tests for other outstanding issues
-- model facets of stream morphology to ascertain if consistent with encounter signatures

% \section{Conclusions}
% \label{sec:conclusions}
